UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2019-05-29 ms/main.tex
more edits

Forced expiratory volume in 1 second (FEV1) (figures~\ref{fig:supp_ukbb_fev_f}, \ref{fig:supp_ukbb_fev_m}) also shows strong correlations with certain populations --- South Asian, African, and Caribbean --- having considerably lower measurements on average (see figures~\ref{fig:supp_box_fev_f} and \ref{fig:supp_box_fev_m} for boxplots and p-values). Notably, there appears to be a juncture in the admixture population, highlighted in figure~\ref{fig:supp_comparison_fev_afr}, where the distribution of FEV1 changes. This roughly corresponds to the transition from Black African/Caribbean individuals to those who identified having mixed backgrounds. Boxplots and statistical testing suggest that relative to White British populations, FEV1 values are significantly lower for Black African and Black Caribbean populations, but not for White and Black Caribbean and White and Black African populations. Unidentified populations highlighted in figure~\ref{fig:supp_comparison_fev_chi_eur} suggest that one ethnic group close to the Chinese may have higher than average FEV1 values compared to the relatively low values of the Chinese themselves; using available country of birth data, most of the unidentified ethnic group were born in Japan. These results merit further investigation and underscore the exploratory value of UMAP --- PCA alone compresses too many populations together to see finer details (figures~\ref{fig:supp_ukbb_pca_height_res_f}, \ref{fig:supp_ukbb_pca_fev_f}), while t-SNE's clustering breaks the desirable continuity between and within populations~(figures~\ref{fig:supp_ukbb_tsne_height_res_f}, \ref{fig:supp_ukbb_tsne_fev_f}).

@@

\subsection*{Discussion}
UMAP comprehensively illustrates genotypic information at fine-scales and within the context of a global population structure. Using it is straightforward and fast; given PCA data and a desktop computer, UMAP can be performed in 15 to 25 minutes on a sample of hundreds of thousands of individuals over tens of dimensions. It excels with larger datasets containing individuals with admixed backgrounds, which present discrete and continuous population structure. By patching together many local genetic neighbourhoods, the visualizations preserve aspects of both local and global structure, unlike any existing method that we are familiar with.

Using UMAP reveals  clusters that would have been difficult to identify via numerous pairwise PCA plots, such as the geographically restricted cluster within the Hispanic population of the HRS, or the splits within the Gujarati and Punjabi population samples in the 1kGP that were not visible through PCA or Admixture analysis alone \sgcomment{How did the Reich paper find it?}\adpcomment{PCA on the Gujarati alone followed by FST. I guess the highlight is it was found without having to look at a sub-population specifically}. Application to the UKBB underscores the strength of UMAP in large cohorts, with fine relationships between genotype and phenotype and geography presented in visualizations that account for natural genetic clustering. Figures~\ref{fig:umap_height_female}, \ref{fig:supp_ukbb_fev_f}, and \ref{fig:supp_ukbb_zoom} demonstrate phenotypic variation within and across clusters, with height showing continuous variation across admixture edges, as expected from genetically controlled traits, and others, such as leukocyte counts or FEV1, showing sharper boundaries, as expected from environmentally determined traits.

\adpcomment{(scrapped a paragraph here)}

@@

We found that pre-processing the data with PCA allowed for considerable time savings. Groupings on ethnicity formed slowly until reaching a stable number around 10 to 15 components. Geographical patterns in the UKBB continued to appear even up to 40 components, as visible in figures~\ref{fig:supp_montage_ukbb_ns} and \ref{fig:supp_montage_ukbb_ew}.

>>>>

Forced expiratory volume in 1 second (FEV1) (figures~\ref{fig:supp_ukbb_fev_f}, \ref{fig:supp_ukbb_fev_m}) also shows strong correlations with certain populations --- South Asian, African, and Caribbean --- having considerably lower measurements on average (see figures~\ref{fig:supp_box_fev_f} and \ref{fig:supp_box_fev_m} for boxplots and p-values). Notably, there appears to be a juncture in the admixture population, highlighted in figure~\ref{fig:supp_comparison_fev_afr}, where the distribution of FEV1 changes. This roughly corresponds to the transition from Black African/Caribbean individuals to those who identified having mixed backgrounds. Boxplots and statistical testing suggest that relative to White British populations, FEV1 values are significantly lower for Black African and Black Caribbean populations, but not for White and Black Caribbean and White and Black African populations. Figure~\ref{fig:supp_comparison_fev_chi_eur} also shows an important difference in FEV1 between self-identified Chinese and a nearby cluster enriched in individuals born in Japan. \sgcomment{boxblot}.

These results merit further investigation and underscore the exploratory value of UMAP --- PCA alone compresses too many populations together to see finer details (figures~\ref{fig:supp_ukbb_pca_height_res_f}, \ref{fig:supp_ukbb_pca_fev_f}), while t-SNE's clustering breaks the desirable continuity between and within populations~(figures~\ref{fig:supp_ukbb_tsne_height_res_f}, \ref{fig:supp_ukbb_tsne_fev_f}).

@@

\subsection*{Discussion}
UMAP comprehensively illustrates genotypic information at fine scales and within the context of global population structure. Using it is straightforward and fast; given PCA data and a desktop computer, UMAP can be performed in 15 to 25 minutes on a sample of hundreds of thousands of individuals over tens of dimensions. It excels with larger datasets containing individuals with admixed backgrounds, which present discrete and continuous population structure. By patching together many local genetic neighbourhoods, the visualizations preserve aspects of both local and global structure, unlike any existing method that we are familiar with.

Using UMAP reveals  clusters that would have been difficult to identify via numerous pairwise PCA plots, such as the geographically restricted cluster within the Hispanic population of the HRS, or the splits within the Gujarati and Punjabi population samples in the 1kGP that were not visible through global PCA or Admixture analysis alone.

%Application to the UKBB underscores the strength of UMAP in large cohorts, with fine relationships between genotype and phenotype and geography presented in visualizations that account for natural genetic clustering. Figures~\ref{fig:umap_height_female}, \ref{fig:supp_ukbb_fev_f}, and \ref{fig:supp_ukbb_zoom} demonstrate phenotypic variation within and across clusters, with height showing continuous variation across admixture edges, as expected from genetically controlled traits, and others, such as leukocyte counts or FEV1, showing sharper boundaries, as expected from environmentally determined traits.

Using UMAP on the UKBB data revealed a range of patterns of covariation between geography, phenotypes and genotypes, with traits such as height showing continuous variation across admixture edges and geographic gradients, as expected from genetically controlled complex traits, and others, such as leukocyte counts or FEV1, showing sharper boundaries and nonlinear behavior consistent with the existence of strong regional environmental influences.

We found that pre-processing the data with PCA allowed for considerable time savings. Groupings on ethnicity formed slowly until reaching a stable number around 10 to 15 components. Geographical patterns in the UKBB continued to appear even up to 40 components, as visible in figures~\ref{fig:supp_montage_ukbb_ns} and \ref{fig:supp_montage_ukbb_ew}. In our experience, we find it useful to include as many PCs as are available.

@@
